Geometric models reveal the hidden structure of conceptual knowledge

### 2023-01-26 paper/main.tex

A core assumption of our modeling framework is that knowledge is ``smooth''
such that knowing about one concept implies knowledge about other related
concepts (i.e., concepts that are nearby in text embedding space). Conversely,
\textit{not} knowing about a given concept implies \textit{less} knowledge
about other related concepts. We characterized how estimated knowledge (and
lack of knowledge) fell off with distance in the text embedding space
(Fig.~\ref{fig:predictions}, top panels).

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} \textbf{Top panels.} Separately for each quiz (panel
    column), we plot the average correlations between the topic vectors of
    held-out correctly (blue) and incorrectly (red) answered questions, and the
    topic vectors for the $n$ closest remaining correctly or incorrectly
    answered questions, respectively. For details, see \textit{Testing the
    ``smoothness'' of knowledge}. \textbf{Bottom panels.} Separately for each
    quiz (panel column), we plot the distributions of predicted knowledge at
    the embedding coordinates of each held out correctly (blue) or incorrectly
    (red) answered question. The $t$-tests reported in each panel are between
    the distributions of estimated knowledge at the coordinates of correctly
    versus incorrectly answered held-out questions. For details, see
    \textit{Estimating held-out conceptual knowledge}.}

@@

For each participant, for each quiz, for each of $c$ correctly answered
question $q$ in turn, we sorted all of the other correctly answered questions
in descending order of their topic vectors' correlations with $q$'s topic
vector. We repeated this for all correctly answered questions to obtain a $c
\times (c - 1)$ matrix of correlations, where the rows corresponded to held-out
questions and the columns corresponded to the most similar (leftmost column),
second most similar (second column), other correctly answered questions. Each
correlation for each pair of questions ($a$ and $b$) appears exactly twice in
this matrix: once when $a$ is the held-out question, and once when $b$ is the
held-out question. We applied the Fisher $z$-transformation to these
correlations, averaged down the rows, and then applied the inverse Fisher
$z$-transformation to obtain a $1 \times (c - 1)$ array of average correlations
for the participant. We repeated this process across all $n$ participants,
yielding an $n \times 14$ matrix (15 is the maximum number of questions that
could have been answered correctly on any single 15-question quiz, so at most
14 correctly answered questions could be compared to a held-out correctly
answered question). Note that some of the entries, for some of the
participants, were empty (e.g., if a given participant answered at least one
question incorrectly, they would never have 14 other correctly answered
questions to compare to a held-out correctly answered question). If a
participant answered \textit{no} questions correctly on a given quiz, their
entire row would be empty.

The blue ribbon plots in the top panels of Figure~\ref{fig:predictions} display
the mean (across participants) and bootstrap-estimated 95\% confidence
intervals of these matrices (i.e., one per quiz, each plotted in a separate
panel). We repeated the same process for \textit{incorrectly} answered
questions (i.e., comparing held-out incorrectly answered questions to other
incorrectly answered questions from the same quiz, for each participant in
turn) to obtain the red ribbon plots in the top panels of
Figure~\ref{fig:predictions}.

The correlation plots for both curves, in all three top panels of
Figure~\ref{fig:predictions}, confirm one aspect of our ``smoothness''
assumption. Specifically, given the topic vector of a correctly answered
question, other correctly answered questions also tend to have similar topic
vectors...

\textbf{JRM NOTE 1: need to think about these results more; stopping here
temporarily to back up progress...}

\textbf{JRM NOTE 2: consider moving ``estimating held-out conceptual
knowledge'' from methods to here.}

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$ topics. We hoped that increasing the number of
topics from 15 to 100 might help us to generalize the knowledge predictions.
(Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.)

@@

@@

\subsubsection*{Estimating held-out conceptual knowledge}

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, the estimated knowledge should have some
predictive information about whether the participant is likely ot answer the
question correctly or incorrectly. For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we aggregated these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and anotehr for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}, bottom
panels). We then used independent samples $t$-tests to compare the means of
these distributions of estimated knowledge.

>>>>

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, the estimated knowledge should have some
predictive information about whether the participant is likely ot answer the
question correctly or incorrectly. For each question in turn, for each
participant, we used Equation~\ref{eqn:prop} to estimate (using all
\textit{other} questions from the same quiz, from the same participant) the
participant's knowledge at the held-out question's embedding coordinate. For
each quiz, we aggregated these estimates into two distributions: one for the
estimated knowledge at the coordinates of each \textit{correctly} answered
question, and another for the estimated knowledge at the coordinates of each
\textit{incorrectly} answered question (Fig.~\ref{fig:predictions}). We then
used independent samples $t$-tests to compare the means of these distributions
of estimated knowledge.

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held out correctly (blue) or incorrectly (red) answered question. The
    $t$-tests reported in each panel are between the distributions of estimated
    knowledge at the coordinates of correctly versus incorrectly answered
    held-out questions.}

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left panel). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered ($t(633) =
0.577, p = 0.564$). After watching the first video, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle panel) exhibited a positive shift relative
to held-out incorrectly answered questions ($t(633) = 3.961, p < 0.001$). After
watching the second video, estimated knowledge (from the third quiz;
Fig.~\ref{fig:predictions}, right panel) for \textit{all} questions exhibited a
positive shift. However, the increase in estimated knowledge for held-out
correctly answered questions was larger than for held-out incorrectly answered
questions ($t(628) = 2.045, p = 0.041$).

Knowledge estimates need not be limited to the content of the lectures. As
illustrated in Figure~\ref{fig:knowledge-maps}, our general approach to
estimating knowledge from a small number of quiz questions may be applied to
\textit{any} content, given its text embedding coordinate. To visualize how
knowledge ``spreads'' through text embedding space to content beyond the
lectures participants watched, we first fit a new topic model to the lectures'
sliding windows with $k = 100$ topics. We hoped that increasing the number of
topics from 15 to 100 might help us to generalize the knowledge predictions.
(Aside from increasing the number of topics from 15 to 100, all other
procedures and model parameters were carried over from the preceding analyses.)
We resampled each lecture's topic trajectory to 1~Hz and also projected each
question into the same 100-dimensional topic space.

@@

We projected the resulting 100-dimensional topic vectors (for each second of
video and for each question) into a shared 2-dimensional space (see
\textit{Creating knowledge and learning map visualizations}). Next, we sampled
points evenly from a $100 \times 100$ grid of coordinates that evenly tiled a
rectangle enclosing the 2D projections of the video trajectories and questions.
We used Equation~\ref{eqn:rbf-knowledge} to estimate participants' knowledge at
each of these 10K sampled locations, and we averaged these estimates across
participants to obtain an estimated average \textit{knowledge map}
(Fig.~\ref{fig:knowledge-maps}). Intuitively, the knowledge map constructed
from a given quiz's responses provides a visualization of how ``much''
participants know about any content expressible by the fitted model.

Several features of the resulting knowledge maps are worth noting. The average
knowledge map estimated from Quiz 1 responses (Fig.~\ref{fig:knowledge-maps},
leftmost map) shows that participants tended to have relatively little
knowledge about any parts of the text embedding space (i.e., the shading is
relatively dark everywhere). The knowledge map estimated from Quiz 2 responses
shows a marked increase in knowledge on the left side of the map (around
roughly the same range of coordinates covered by the \textit{Four Fundamental
Forces} lecture, indicated by the dotted blue line). In other words,
participants' estimated increase in knowledge is localized to conceptual
content that is nearby (i.e., related to) the content from the lecture they
watched prior to taking Quiz 2. This localization is non-trivial: the knowledge
estimates are informed only by the embedded coordinates of the \textit{quiz
questions}, not by the embeddings of either lecture. Finally, the knowledge map
estimated from Quiz 3 responses shows a second increase in knowledge, localized
to the region surrounding the embedding of the \textit{Birth of Stars} lecture
participants watched prior to taken Quiz 3.

Another way of visualizing these content-specific increases in knowledge
(apparently driven by watching each lecture) is displayed in
Figure~\ref{fig:knowledge-maps}B. Taking the point-by-point difference between
the knowledge maps estimated from responses to a successive pair of quizzes
yields a \textit{learning map} that describes the \textit{change} in knowledge
estimates from one quiz to the next. These learning maps highlight that the
estimated knowledge increases we observed across maps were specific to the
regions around the embeddings of each lecture in turn.

Because the 2D projection we used to construct the knowledge and learning maps
is (partially) invertable, we may gain additional insights into the estimates
by reconstructing the original high-dimensional topic vectors for any point(s)
in the maps we are interested in. For example, this could serve as a useful
tool for an instructor looking to better understand which content areas a
student (or a group of students) knows well (or poorly). As a demonstration, we
show the top-weighted words from the blends of topics reconstructed from three
locations on the maps (Fig.~\ref{fig:knowledge-maps}C): one point near the
\textit{Four Fundamental Forces} embedding (yellow); a second point near the
\textit{Birth of Stars} embedding (orange), and a third point somewhere in
between the two lectures' embeddings (pink). As shown in the word clouds in the
Panel, the top-weighted words at the coordinate near the \textit{Four
Fundamental Forces} embedding also tended to be weighted heavily by the topics
expressed in that lecture. Similarly, the top-weighted words at the coordinate
near the \textit{Birth of Stars} embedding tended to be weighted most heavily
by the topics expressed in \textit{that} lecture. And the top-weighted words at
the coordinate between the two lectures' embeddings show a roughly even mix of
words most strongly associated with each lecture.

@@
